Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2025-05-25 paper/main.tex
add a few more changes to main text

\begin{abstract}\noindent We develop a mathematical framework, based on natural
language processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high-dimensional
representation space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
answered small sets of multiple-choice quiz questions interleaved between
watching two course videos from the Khan Academy platform. We apply our
framework to the videos' transcripts and the text of the quiz questions to
quantify the content of each moment of video and each quiz question. We use
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video and predict their success
on individual quiz questions. Our findings show how a small set of quiz
questions may be used to obtain rich and meaningful high-resolution insights
into what each learner knows, and how their knowledge changes over time as they
learn.

@@

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should
carry some predictive information about whether they are likely to answer that
question correctly or incorrectly. We developed a statistical approach to test
this claim. For each quiz question a participant answered, in turn, we used
Equation~\ref{eqn:prop} to estimate their knowledge at the given question's
embedding-space coordinate based on other questions that participant answered
on the same quiz. We repeated this for all participants, and for each of the
three quizzes. Then, separately for each quiz, we fit a generalized linear
mixed model (GLMM) with a logistic link function to explain the probability of
correctly answering a question as a function of estimated knowledge at its
embedding coordinate, while accounting for varied effects of individual
participants and questions (see \nameref{subsec:glmm}). To assess the predictive
value of the knowledge estimates, we compared each GLMM to an analogous (i.e.,
nested) ``null'' model that assumed these estimates carried no predictive information using parametric bootstrap likelihood-ratio tests.

@@

We carried out three different versions of the analyses described above,
wherein we considered different sources of information in our estimates of
participants' knowledge for each quiz question. First, we estimated knowledge
at each question's embedding coordinate using \textit{all} other questions
answered by the same participant on the same quiz (``All questions'';
Fig.~\ref{fig:predictions}, top row). This test was intended to assess the
overall predictive power of our approach. Second, we estimated knowledge for
each question about a given lecture using only the other questions (from the
same participant and quiz) about that \textit{same} lecture
(``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows). This test was
intended to assess the \textit{specificity} of our approach by asking whether
our predictions could distinguish between questions about different content
covered by the same lecture. Third, we estimated knowledge for each question
about one lecture using only the questions (from the same participant and quiz)
about the \textit{other} lecture (``Across-lecture'';
Fig.~\ref{fig:predictions}, bottom rows). This test was intended to assess the
\textit{generalizability} of our approach by asking whether our predictions
could extend across the content areas of the two lectures. When estimating
participants' knowledge, we used a rebalancing procedure to ensure that (for a
given participant and quiz) their knowledge estimates for correctly and
incorrectly answered questions were computed from the same underlying proportion
of correctly answered questions (see~\nameref{subsec:glmm}).

@@

We observed a similar set of results when we restricted our estimates of
participants' knowledge for questions about each lecture to consider only
their performance on other questions about the \textit{same} lecture.
Specifically, for Quiz~1, participants' knowledge of \textit{Four Fundamental
Forces}-related questions, estimated from their performance on other
\textit{Four Fundamental Forces}-related questions, was predictive of their
ability to answer those questions correctly ($OR = 15.934,\ 95\%\
\textnormal{CI} = [5.173,\ 38.005],\ \lambda_{LR} = 40.971,\ p = 0.001$). The
same was true of participants' estimated knowledge for \textit{Birth of
Stars}-related questions based on their performance on other \textit{Birth of
Stars}-related questions ($OR = 9.775,\ 95\%\ \textnormal{CI} = [2.93,\
25.08],\ \lambda_{LR} = 13.924,\ p = 0.001$). These within-lecture knowledge
estimates also predicted success on questions about both lectures when we
computed them analogously for Quiz~2 (\textit{Four Fundamental Forces}: $OR =
35.126,\ 95\%\ \textnormal{CI} = [5.113,\ 123.868],\ \lambda_{LR} = 32.251,\
p < 0.001$; \textit{Birth of Stars}: $OR = 4.717,\ 95\%\ \textnormal{CI} =
[2.021,\ 9.844],\ \lambda_{LR} = 16.788,\ p < 0.001$). For Quiz~3, we found that
within-lecture knowledge estimates predicted participants' success on
\textit{Birth of Stars}-related questions ($OR = 16.902,\ 95\%\ \textnormal{CI}
= [3.353,\ 53.265],\ \lambda_{LR} = 23.233,\ p < 0.001$) but not on \textit{Four
Fundamental Forces}-related questions ($OR = 2.485,\ 95\%\ \textnormal{CI} =
[0.724,\ 8.366],\ \lambda_{LR} = 1.984,\ p = 0.170$). This may indicate that the
within-lecture knowledge estimates are susceptible to ceiling effects in
participants' quiz performance. On Quiz~3, after viewing both lectures, no
participant answered more than three \textit{Four Fundamental Forces}-related
questions incorrectly, and all but five participants (out of 50) answered two or
fewer incorrectly. (This was the only subset of questions about either lecture, 
across all three quizzes, for which this was true.) Because of this, for 90\% of 
participants, our within-lecture estimates of their knowledge for \textit{Four 
Fundamental Forces}-related questions that they answered incorrectly leveraged 
information from at most a single other question they were \textit{not} able to 
correctly answer. This likely hampered our ability to accurately characterize 
the specific (and by the time they took Quiz~3, relatively few) aspects of the 
lecture content these participants did \textit{not} know about, and successfully 
distinguish them from the far more numerous aspects of the lecture content they 
now \textit{did} know about. Taken together, these results suggest that our 
knowledge estimates can reliably distinguish between questions about different
content covered by a single lecture, provided there is sufficient diversity in 
participants' quiz responses to extract meaningful information about both what 
they know and what they do not know.

Finally, we estimated participants' knowledge for each question about one
lecture using their performance on questions (from the same quiz) about the
\textit{other} lecture. This is an especially stringent test of our approach.
Our primary assumption in constructing our knowledge estimates is that knowledge
about a given concept is similar to knowledge about other concepts that are
nearby in the embedding space. However, our analyses in Figure~\ref{fig:topics}
and Supplementary Figure~\topicWeights\ show that the embeddings of content from
the two lectures (and of their associated quiz questions) are largely distinct 
from each other. Therefore, any predictive power of these across-lecture 
knowledge estimates must overcome large distances in the embedding space. To put 
this in concrete terms, this test requires predicting participants' performance 
on individual, highly specific questions about the formation of stars from their 
responses to just five multiple-choice questions about the fundamental forces of 
the universe (and vice versa).

@@

That the knowledge predictions derived from the text embedding space reliably
distinguish between correctly and incorrectly answered held-out questions
(Fig.~\ref{fig:predictions}) suggests that spatial relationships within this
space can help explain what participants know. But how far does this
explanatory power extend? For example, suppose we know that a participant
correctly answered a question at embedding coordinate $x$. As we move farther
away from $x$ in the embedding space, how does the likelihood that the
participant knows about the content at a given location ``fall off'' with
distance? Conversely, suppose the participant instead answered that same
question \textit{incorrectly}. Again, as we move farther away from $x$ in the
embedding space, how does the likelihood that the participant does \textit{not}
know about the content at a given coordinate change with distance? We reasoned that,
assuming our embedding space is capturing something about how individuals
actually organize their knowledge, a participant's ability to answer questions
embedded very close to $x$ should tend to be similar to their ability to answer
the question embedded \textit{at} $x$. But once we
reach some sufficiently large distance from $x$, our ability to infer whether
or not a participant will correctly answer a question based on their ability to
answer the question at $x$ should be no better than guessing based on their
\textit{overall} proportion of correctly answered questions. In other words,
beyond the maximum distance at which a participant's ability to answer the
question at $x$ is informative of their ability to answer a second question at
location $y$, guessing the outcome at $y$ based on the outcome at $x$ should be no more
successful than guessing based on a measure that does not consider
embedding-space distance.

@@

In our study, we characterize the ``coordinates'' of participants' knowledge
using a relatively simple ``bag-of-words'' text embedding model~\citep[LDA;
][]{BleiEtal03}. More sophisticated text embedding models, such as
transformer-based models~\citep{ViswEtal17, DevlEtal18, ChatGPT, TouvEtal23},
can leverage additional textual information such as complex grammatical and 
semantic relationships between words, higher-order syntactic structures, 
stylistic features, and more. We considered using transformer-based models in 
our study, but we found that the text embeddings derived from these models were 
surprisingly uninformative with respect to differentiating or otherwise 
characterizing the conceptual content of the lectures and questions we used 
(see \suppResults). We suspect that this reflects a broader challenge in 
constructing models that are both high-resolution within a given domain (e.g., 
the domain of physics lectures and questions) \textit{and} sufficiently broad 
as to enable them to cover a wide range of domains. Essentially, these 
``larger'' language models learn these more complex features of language through 
training on enormous and diverse text corpora. But as a result, their 
embedding spaces also ``span'' an enormous and diverse range of conceptual 
content, sacrificing a degree of specificity in their capacities to distinguish 
subtle conceptual differences within a more narrow range of content. 
In comparing our LDA model (trained specifically on the lectures used in our 
study) to a larger transformer-based model (BERT), we found that our LDA model provides 
both coverage of the requisite material and specificity at the level of 
individual questions, while BERT essentially relegates the contents of both 
lectures and all quiz questions (which are all broadly about ``physics'') to a 
tiny region of its embedding space, thereby blurring out meaningful distinctions 
between different specific concepts covered by the lectures and questions 
(Supp.~Fig.~\ldaVsBERT). We note that these are not criticisms of BERT, nor of 
other large language models trained on large and diverse corpora. Rather, our 
point is that simpler models trained on relatively small but specialized 
corpora can outperform much more complex models trained on much larger corpora 
when we are specifically interested in capturing subtle conceptual differences 
at the level of a single course lecture or quiz question. On the other hand, if 
our goal had been to choose a model that generalized to many different content 
areas simultaneously, we would expect our LDA model to perform comparatively poorly to BERT or 
other much larger general-purpose models. We suggest that bridging this tradeoff 
between high resolution within a single content area and the ability to 
generalize to many diverse content areas will be an important challenge for 
future work.

At the opposite end of the spectrum from large language models, one could also
imagine using an even \textit{simpler} ``model'' than LDA that relates the 
contents of course lectures and quiz questions through explicit word-overlap 
metrics (rather than similarities in the latent topics they exhibit). In a 
supplementary analysis (Supp.~Fig.~\jaccard), we compared the LDA-based 
question-lecture matches shown in Figure~\ref{fig:question-correlations} with
analogous matches based on the Jaccard similarity between each question's text 
and each sliding window from the corresponding lecture's transcript. Similarly 
to the embeddings derived from BERT, we found that this approach also blurred 
meaningful distinctions between concepts presented in different parts of each 
lecture and tested by different quiz questions. But rather than characterizing
their contents at too \textit{broad} a semantic scale, the lack of specificity 
in this approach arises from considering too \textit{narrow} a semantic scale: 
the sorts of concepts typically conveyed in course lectures and tested by quiz 
questions are not defined (and meaningful similarities and distinctions between 
them do not tend to emerge) at the level of individual words.

>>>>

\begin{abstract}\noindent We develop a mathematical framework, based on natural
language processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high-dimensional
representation space where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
answered small sets of multiple-choice quiz questions interleaved between
watching two Khan Academy course videos. We apply our
framework to the videos' transcripts and the text of the quiz questions to
quantify the content of each moment of video and each quiz question. We use
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video and predict their success
on individual quiz questions. Our findings show how a small set of quiz
questions may be used to obtain rich and meaningful high-resolution insights
into what each learner knows, and how their knowledge changes over time as they
learn.

@@

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should
carry some predictive information about whether they are likely to answer that
question correctly or incorrectly. We developed a statistical approach to test
this claim. For each quiz question a participant answered, in turn, we used
Equation~\ref{eqn:prop} to estimate their knowledge at that question's
embedding-space coordinate based on other questions that participant answered
on the same quiz. We repeated this for all participants, and for each of the
three quizzes. Then, separately for each quiz, we fit a generalized linear
mixed model (GLMM) with a logistic link function to explain the probability of
correctly answering a question as a function of estimated knowledge at its
embedding coordinate, while accounting for varied effects of individual
participants and questions (see \nameref{subsec:glmm}). To assess the predictive
value of the knowledge estimates, we compared each GLMM to an analogous (i.e.,
nested) ``null'' model that assumed these estimates carried no predictive information using parametric bootstrap likelihood-ratio tests.

@@

We carried out three different versions of the analyses described above,
wherein we considered different sources of information in our estimates of
participants' knowledge for each quiz question. First, we estimated knowledge
at each held-out question's embedding coordinate using \textit{all} other questions
answered by the same participant on the same quiz (``All questions'';
Fig.~\ref{fig:predictions}, top row). This test was intended to assess the
overall predictive power of our approach. Second, we estimated knowledge for
each question about a given lecture using only the other questions (from the
same participant and quiz) about that \textit{same} lecture
(``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows). This test was
intended to assess the \textit{specificity} of our approach by asking whether
our predictions could distinguish between questions about different content
covered by the same lecture. Third, we estimated knowledge for each question
about one lecture using only the questions (from the same participant and quiz)
about the \textit{other} lecture (``Across-lecture'';
Fig.~\ref{fig:predictions}, bottom rows). This test was intended to assess the
\textit{generalizability} of our approach by asking whether our predictions
could extend across the content areas of the two lectures. When estimating
participants' knowledge, we used a rebalancing procedure to ensure that (for a
given participant and quiz) their knowledge estimates for correctly and
incorrectly answered questions were computed from the same underlying proportion
of correctly answered questions (see~\nameref{subsec:glmm}).

@@

We observed a similar set of results when we restricted our estimates of
participants' knowledge to consider only
their performance on other questions about the \textit{same} lecture.
Specifically, for Quiz~1, participants' knowledge of \textit{Four Fundamental
Forces}-related questions, estimated from their performance on other
\textit{Four Fundamental Forces}-related questions, was predictive of their
ability to answer those questions correctly ($OR = 15.934,\ 95\%\
\textnormal{CI} = [5.173,\ 38.005],\ \lambda_{LR} = 40.971,\ p = 0.001$). The
same was true of participants' estimated knowledge for \textit{Birth of
Stars}-related questions based on their performance on other \textit{Birth of
Stars}-related questions ($OR = 9.775,\ 95\%\ \textnormal{CI} = [2.93,\
25.08],\ \lambda_{LR} = 13.924,\ p = 0.001$). These results also held for
participants' Quiz~2 responses (\textit{Four Fundamental Forces}: $OR =
35.126,\ 95\%\ \textnormal{CI} = [5.113,\ 123.868],\ \lambda_{LR} = 32.251,\
p < 0.001$; \textit{Birth of Stars}: $OR = 4.717,\ 95\%\ \textnormal{CI} =
[2.021,\ 9.844],\ \lambda_{LR} = 16.788,\ p < 0.001$) and partially for 
their Quiz~3 responses (\textit{Birth of Stars}: $OR = 16.902,\ 95\%\ 
\textnormal{CI} = [3.353,\ 53.265],\ \lambda_{LR} = 23.233,\ p < 0.001$; 
\textit{Four Fundamental Forces}: $OR = 2.485,\ 95\%\ \textnormal{CI} =
[0.724,\ 8.366],\ \lambda_{LR} = 1.984,\ p = 0.170$). Speculatively, the Quiz~3 
results suggest that the within-lecture knowledge estimates may be 
susceptible to ceiling effects in
participants' quiz performance. On Quiz~3, after viewing both lectures, no
participant answered more than three \textit{Four Fundamental Forces}-related
questions incorrectly, and all but five participants (out of 50) answered two or
fewer incorrectly. (This was the only subset of questions about either lecture, 
across all three quizzes, for which this was true.) Consequently, for 90\% of 
participants, our within-lecture estimates of their knowledge for \textit{Four 
Fundamental Forces}-related questions that they answered incorrectly leveraged 
information from at most a single other question they were \textit{not} able to 
correctly answer. This likely hampered our ability to accurately characterize 
the specific (and by the time they took Quiz~3, relatively few) aspects of the 
lecture content these participants did \textit{not} know about, and successfully 
distinguish them from the far more numerous aspects of the lecture content they 
now \textit{did} know about. Taken together, these within-lecture results suggest that our 
knowledge estimates can reliably distinguish between questions about different
content covered by a single lecture, provided there is sufficient diversity in 
participants' quiz responses to extract meaningful information about both what 
they know and what they do not know.

Finally, we estimated participants' knowledge for each question about each
lecture using only their performance on questions (from the same quiz) about the
\textit{other} lecture. This is an especially stringent test of our approach.
Our primary assumption in constructing our knowledge estimates is that knowledge
about a given concept is similar to knowledge about other concepts that are
nearby in the embedding space. However, our analyses in Figure~\ref{fig:topics}
and Supplementary Figure~\topicWeights\ show that the embeddings of content from
the two lectures (and of their associated quiz questions) are largely distinct 
from each other. Therefore, any predictive power of these across-lecture 
knowledge estimates must overcome large distances in the embedding space. To put 
this in concrete terms, this test requires predicting participants' performance 
on individual, highly specific questions about the formation of stars from their 
responses to just five multiple-choice questions about the fundamental forces of 
the universe (and vice versa).

@@

That the knowledge predictions derived from the text embedding space reliably
distinguish between correctly and incorrectly answered held-out questions
(Fig.~\ref{fig:predictions}) suggests that geometric relationships within this
space can help explain what participants know. But how far does this
explanatory power extend? For example, suppose we know that a participant
correctly answered a question at embedding coordinate $x$. As we move farther
away from $x$ in the embedding space, how does the likelihood that the
participant knows about the content at a given location ``fall off'' with
distance? Conversely, suppose the participant instead answered that same
question \textit{incorrectly}. Again, as we move farther away from $x$ in the
embedding space, how does the likelihood that the participant does \textit{not}
know about the content at a given coordinate change with distance? We reasoned that,
assuming our embedding space is capturing something about how individuals
actually organize their knowledge, a participant's ability to answer questions
embedded very close to $x$ should tend to be similar to their ability to answer
the question embedded \textit{at} $x$. But once we
reach some sufficiently large distance from $x$, our ability to infer whether
or not a participant will correctly answer a question based on their ability to
answer the question at $x$ should be no better than guessing based on their
\textit{overall} proportion of correctly answered questions. In other words,
beyond the maximum distance at which a participant's ability to answer the
question at $x$ is informative of their ability to answer a second question at
location $y$, guessing the outcome at $y$ based on the outcome at $x$ should be no more
successful than guessing based on a measure that does not consider
embedding-space distance.

@@

In our study, we characterize the ``coordinates'' of participants' knowledge
using a relatively simple ``bag-of-words'' text embedding model~\citep[LDA;
][]{BleiEtal03}. More sophisticated text embedding models, such as
transformer-based models~\citep{ViswEtal17, DevlEtal18, ChatGPT, TouvEtal23},
can leverage additional textual information such as complex grammatical and 
semantic relationships between words, higher-order syntactic structures, 
stylistic features, and more. We considered using transformer-based models in 
our study, but we found that the text embeddings derived from these models were 
surprisingly uninformative with respect to differentiating or otherwise 
characterizing the conceptual content of the lectures and questions we used 
(see \suppResults). We suspect that this reflects a broader challenge in 
constructing models that are both high-resolution within a given domain (e.g., 
the domain of physics lectures and questions) \textit{and} sufficiently broad 
as to enable them to cover a wide range of domains. Essentially, 
``larger'' language models learn more complex features of language through 
training on enormous and diverse text corpora. But as a result, their 
embedding spaces also ``span'' an enormous and diverse range of conceptual 
content, sacrificing a degree of specificity in their capacities to distinguish 
subtle conceptual differences within a more narrow range of content. 
In comparing our LDA model (trained specifically on the lectures used in our 
study) to a larger transformer-based model (BERT), we found that our LDA model provides 
both coverage of the requisite material and specificity at the level of 
individual questions, while BERT essentially relegates the contents of both 
lectures and all quiz questions (which are all broadly about ``physics'') to a 
tiny region of its embedding space, thereby blurring out meaningful distinctions 
between different specific concepts covered by the lectures and questions 
(Supp.~Fig.~\ldaVsBERT). We note that these are not criticisms of BERT, nor of 
other large language models trained on large and diverse corpora. Rather, our 
point is that simpler models trained on relatively small but specialized 
corpora can outperform much more complex models trained on much larger corpora 
when we are specifically interested in capturing subtle conceptual differences 
at the level of a single, narrowly focused course lecture or quiz question. On the other hand, if 
our goal had been to choose a model that generalized to many different content 
areas simultaneously, we would expect our LDA model to perform comparatively poorly to BERT or 
other much larger general-purpose models. We suggest that bridging this tradeoff 
between achieving high resolution within a single content area and the ability to 
generalize to many diverse content areas will be an important challenge for 
future work.

At the opposite end of the spectrum from large language models, one could also
imagine using an even \textit{simpler} ``model'' than LDA that relates the 
contents of course lectures and quiz questions through explicit word-overlap 
metrics (rather than similarities in the latent topics they exhibit). In a 
supplementary analysis (Supp.~Fig.~\jaccard), we compared the LDA-based 
question-lecture matches shown in Figure~\ref{fig:question-correlations} with
analogous matches based on the Jaccard similarity between each question's text 
and each sliding window from the corresponding lecture's transcript. As for 
the embeddings derived from BERT, we found that this word-matching approach also blurred 
meaningful distinctions between concepts presented in different parts of each 
lecture and tested by different quiz questions. But rather than characterizing
their contents at too \textit{broad} a semantic scale, the lack of specificity 
in this approach arises from considering too \textit{narrow} a semantic scale: 
the sorts of concepts typically conveyed in course lectures and tested by quiz 
questions are not defined (and meaningful similarities and distinctions between 
them do not tend to emerge) at the level of individual words.
